---
title: "Homework 2: Visualizing data"
author: "Andrés Lajous"
date: "Due: Sunday, Feb 15, 2015"
output: html_document
---

Topics covered in this homework include:

- data tidying
- dplyr 
- ggplot2


### Let's return to Prof. Rodriguez's final exam from last semester.  Now you will redo some of the same problems using dplyr and ggplot2.

From your exam: "The data fraem Robey in the `car` package has data on fertility and contraceptive use in 50 developing countries around 1990.  There are three variables (1) region, a factor with four levels (Africa, Asia, Latin America, and Near East) (2) tfr, the total fertiility rate, the average number of children that a women would have over her lifetime at current rates (usually abbreviated TFR), and (3) contraceptors, the precentage of contraception among women of childbearing age."

(@) **"Let us say that a country has undergone a contraceptive revolution when at least half of the women in childbearing age use contraception.  Calculate and plot the proportion of countries that have achived this milestone by region."  Be sure to use dplyr and ggplot2.**

```{r}
library(ggplot2)
library(dplyr)
library(car)
Robey <- tbl_df(Robey)

## get the total of countries by region
total <- Robey %>%
  group_by(region)%>%
  summarise(total = n())

## get the total of countries that have gone through a contraceptive revolution by region
sup <- Robey %>% 
  filter(contraceptors >= 50) %>%
  group_by(region)%>%
  summarise(sup = n())
  
## join the twto tables and create a vector with the proportion of countrie per region
prop <- left_join(total, sup, by = "region")  %>%
  mutate(prop = sup/total)

## plot with gg plot. add y lable, and title
plot1 <- ggplot(data=prop, aes(x=region, y=prop))
plot1 + geom_point(size = 4, color = "blue") +
  ylab("proportion") + 
  ggtitle("Proportion of countries with contraceptive revolution\n by region")
```

(@) **"Compute the average TFR in each region and plot your results."  Be sure to use dplyr and ggplot2.**

```{r}
## create table with mean tfr grouped by region
avg.tfr <- Robey %>%
  group_by(region) %>%
  summarise (avg = mean(tfr))

## plot mean tfr by region
plot2 <- ggplot(data = avg.tfr, aes(x = region, y = avg))
plot2 + geom_point(size = 4, color = "blue") +
  ylab("tfr") + 
  ggtitle("Average TFR by Region")
```

(@) **"Plot the TFR against the proportion of married women using contraception and superimpose a regression line." **

```{r}
plot3 <- ggplot(data = Robey, aes(x = contraceptors, y =  tfr))
plot3 + geom_point(size = 2, color = "orange") + 
  layer(stat="smooth", stat_params=list(method="lm")) +
  xlab("percentage of contraception among women") + ylab("total fertility rate") +
  ggtitle("Relationship between contraception among women \n and fertility rate")
```

(@) **Next, create the same plot but with 4 factets, one for each region. **

```{r}
## create the same plot as the previous one, and added fecet grid by region
plot4 <- ggplot(data = Robey, aes(x = contraceptors, y =  tfr))
plot4 + geom_point(size = 2, color = "orange") + 
  layer(stat="smooth", stat_params=list(method="lm")) +
  xlab("percentage of contraception among women") + ylab("total fertility rate") +
  ggtitle("Relationship between contraception among women
          and fertility rate by region") +
  facet_grid(region ~ .)
```

(@) **In words, what would you conclude from the graph above. **

```{answer}
I would conclude that there is a negative relationship between the total fertiltiy rate an the use of contraception among women, regardless of the region of the world where they are localted. 
```

### Next you are going to reproduce and extend this graph by Brad Boehmke that shows median income growth rate by state.

<src img="images/median_income_trends.png">

You can read about it and see all of his code here: https://rpubs.com/bradleyboehmke/median_income_trends

The start code below gets you to where Brad Boehmke's code begins.  

```{r echo=FALSE}
# Preprocessing & summarizing data
library(dplyr)
library(tidyr)

# Visualizatin development
library(grid)
library(scales)
library(ggplot2)

library(xlsx)
# h08.xls is the spreadsheet that was downlaoded from the census website
# https://www.census.gov/hhes/www/income/data/historical/household/2013/h08.xls


data <- read.xlsx("data/h08.xls", sheetName="h08",
                  startRow=62, endRow=113, header=FALSE)

# remove all standard error columns
se.columns <- seq(from=3, to=61, by=2)
data <- select(data, -se.columns)

# check size
if (ncol(data) != 31) {
  # state + 30 years (1984 - 2013)
  stop("ERROR: ncol(data) != 31")
}
if (nrow(data) != 52) {
  # 50 states + plus DC + US
  stop("ERROR: nrow(data) != 52")
}

# create column names
years <- seq(from = 2013, to = 1984)
years.with.x <- paste("X", years, sep="")
c.names <- c("State", years.with.x)
colnames(data) <- c.names

# check that data is the same as the data file he is using
print(data)
```

(@) Write code to reproduce the graph above.  

A few notes:

- In the past we have recommended using `tbl_df()` but we don't recommend that here because it will conflict with some of the way Boehmke wrote his code.  If you want to use tbl_df() you will need to rewrite some of Boehmke's code.

- In step 7 of the code, you should do as much as possible with ggplot2.  Do not use illustrator or any other program.  We want to avoid things that are not automated.


```{r}
us <- data %>%
  filter(State == "United States") %>%
  gather(Year, Income, X2013:X1984) %>%
  separate(Year, c("left", "Year"), sep = "X") %>%
  select(-left) %>%
  arrange(Year)

us_base <- us[us$Year == 1984, 3]
us$us_baseline <- us_base

us <- us %>% 
  mutate(us_change = (Income - us_baseline) / us_baseline)

states <- data %>%
  filter(State != "United States") %>%
  gather(Year, Income, X2013:X1984) %>%
  separate(Year, c("left", "Year"), sep = "X") %>%
  select(-left) %>%
  arrange(Year) %>%
  filter(Income != "NA")

state_base <- states %>%
  filter(Year == 1984)  %>%
  select(State, State_Baseline = Income)

states <- states  %>%
  left_join(state_base)  %>%
  arrange(State)  %>%
  mutate(state_change = (Income - State_Baseline) / State_Baseline)

states$Year <- as.numeric(states$Year)
us$Year <- as.numeric(us$Year)

rank <- states %>%
  filter(Year == 2013)  %>%
  arrange(desc(state_change))  %>%
  mutate(rank = seq(1, length(State), by = 1)) %>%
  filter(rank < 6 | rank > 46)

## create labels for top 5 and low 5
labels <- paste(c(" ---- D.C.", " ---- SD", " ---- IA", " ---- NH", " ---- UT", " ---- DE",
                  " ---- OH", "---- NC"," ---- AK", " ---- NV"), 
               (round(rank$state_change * 100, digits = 1)), c("%"))


## tried to do it with dplyr, but couldn't make it work
##names <- rank%>%
##  mutate(labels = paste(c(" --- D.C.", " ---- SD", " -- IA", " ---- NH", " -- UT", " -- DE",
##                        " ---- OH","-- NC"," ---- AK", " -- NV"), 
##               (round(state_change * 100, digits = 1)))) 
  

p <- ggplot(states, aes(Year, state_change, group = State)) +
  theme_bw() +
  theme(plot.background = element_blank(),
        panel.grid.minor = element_blank(), 
        panel.grid.major.x = element_blank(),
        panel.grid.major.y = element_line(linetype = 3, colour = "grey50"), 
        panel.border = element_blank(), 
        panel.background = element_blank(),
        axis.ticks = element_blank(),  
        axis.title = element_blank()) + 
  geom_line(colour = "grey90", alpha = .9)

p <- p + 
  geom_line(data = us, aes(Year, us_change, group = 1), linetype = 5)

p <- p + 
  geom_line(data = filter(states, State == "Ohio"),
            aes(Year, state_change, group = State), colour = "dodgerblue", size = 1)

p <- p + 
  geom_line(data = filter(states, State == "D.C."),
            aes(Year, state_change, group = State), colour = "grey70") +
  geom_line(data = filter(states, State == "Nevada"),
            aes(Year, state_change, group = State), colour = "grey70")

p <- p + 
  geom_point(data = rank, aes(Year, state_change), shape = 21, size = 1.5, alpha = .6) +
  geom_point(data = filter(us, Year == 2013), aes(Year, us_change), size = 2.5, alpha = .6)


p <- p +
  scale_y_continuous(limits = c(-.2, .55), breaks = seq(-.2, .4, by = .2), label = percent) +
  scale_x_continuous(limits = c(1983, 2017), breaks = seq(1985, 2010, by = 5), expand = c(0, .25))

p <- p +
  annotate("text", x = 1986, y = .42, label = "Change Since 1984", size = 4 , family = "Times")

## add labels for lowest and highest ranking, and also for overall U.S.
p <- p + 
  geom_text(data = rank, aes(Year + 2, state_change), label = labels, size = 3, family = "Times") +
  annotate("text", x = 2015, y = .085, size = 3, family = "Times", label = "Overall U.S. 8.5 %", fontface = "bold")

##create subtitle

subtitle <- "Median income growth since 1984 peaked in 1999 at 18.9% and
has since declined to 8.5% in 2013*. This chart shows the top 5 growth states along
with the bottom 5 contracting states...which, unfortunately, includes my own state of Ohio."

##create footnotes, this is ugly but I wasn't able to align text on the right side of the plot

rfoot <- "\n*Underlying data represents median houshold income                       Source: U.S. Census Bureau | Bradley Boehmke 
adjusted for inflation to 2013 base year dollars. 2014 data                           January 22, 2015
is not yet available."

##add title, subtitles and footnotes with formatting. I was not able to add footnoes unless, 
## I removed the breaks every five years and went back to the default
p <- p + 
  ggtitle("Median Income Growth (1984 - 2013)") + 
 annotate("text", x = 1983, y = .55, label = subtitle,
          family = "Times", size = 4, hjust = 0) + 
  labs(x = rfoot, y ="") +
  theme(plot.title = element_text(hjust = 0, vjust = 1, family = "Times", size = 16), 
        axis.title.x = element_text(family = "Times", colour = "grey70", size = 10, hjust = 0))



```

(@) Modify or extend the graph somehow.

```{r}

##get the average change by state
avg_state <- states %>%
  group_by(State) %>%
  summarise(avg_change = mean(state_change)) %>%
  arrange(avg_change) # lowest one is Kansas


p <- ggplot(states, aes(Year, state_change, group = State)) +
  theme_bw() +
  theme(plot.background = element_blank(),
        panel.grid.minor = element_blank(), 
        panel.grid.major.x = element_blank(),
        panel.grid.major.y = element_line(linetype = 3, colour = "grey50"), 
        panel.border = element_blank(), 
        panel.background = element_blank(),
        axis.ticks = element_blank(),  
        axis.title = element_blank()) + 
  geom_line(colour = "grey90", alpha = .9)

p <- p + 
  geom_line(data = us, aes(Year, us_change, group = 1), linetype = 5)

p <- p + 
  geom_line(data = filter(states, State == "Ohio"),
            aes(Year, state_change, group = State), colour = "dodgerblue", size = 1)

##add red line on Kansas to mark it as the one with the lowest  income growth

p <- p + 
  geom_line(data = filter(states, State == "Kansas"),
            aes(Year, state_change, group = State), colour = "red", size = 1)

p <- p + 
  geom_line(data = filter(states, State == "D.C."),
            aes(Year, state_change, group = State), colour = "grey70") +
  geom_line(data = filter(states, State == "Nevada"),
            aes(Year, state_change, group = State), colour = "grey70")
## add point to kansas
p <- p + 
  geom_point(data = rank, aes(Year, state_change), shape = 21, size = 1.5, alpha = .6) +
  geom_point(data = filter(us, Year == 2013), aes(Year, us_change), size = 2.5, alpha = .6) +
  geom_point(data = filter(states, Year == 2013 & State == "Kansas"), aes(Year, state_change), size = 2.5, alpha = .6) 

 

p <- p +
  scale_y_continuous(limits = c(-.2, .55), breaks = seq(-.2, .4, by = .2), label = percent) +
  scale_x_continuous(limits = c(1983, 2017), breaks = seq(1985, 2010, by = 5), expand = c(0, .25))

p <- p +
  annotate("text", x = 1986, y = .42, label = "Change Since 1984", size = 4 , family = "Times")

## add labels for lowest and highest ranking, and also for overall U.S.
p <- p + 
  geom_text(data = rank, aes(Year + 2, state_change), label = labels, size = 3, family = "Times") +
  annotate("text", x = 2015, y = .085, size = 3, family = "Times", label = "Overall U.S. 8.5 %", fontface = "bold") +
  annotate("text", x = 2015, y = -.021, size = 3, family = "Times", label = "Kansas -2.1 %")

##create subtitle

subtitle <- "Median income growth since 1984 peaked in 1999 at 18.9% and
has since declined to 8.5% in 2013*. This chart shows the top 5 growth states along
with the bottom 5 contracting states...which, unfortunately, includes my own state of Ohio."

##create footnotes, this is ugly but I wasn't able to align text on the right side of the plot

rfoot <- "\n*Underlying data represents median houshold income                       Source: U.S. Census Bureau | Bradley Boehmke 
adjusted for inflation to 2013 base year dollars. 2014 data                           January 22, 2015
is not yet available."

##add title, subtitles and footnotes with formatting. I was not able to add footnoes unless, 
## I removed the breaks every five years and went back to the default
p <- p + 
  ggtitle("Median Income Growth (1984 - 2013)") + 
 annotate("text", x = 1983, y = .55, label = subtitle,
          family = "Times", size = 4, hjust = 0) + 
  labs(x = rfoot, y ="") +
  theme(plot.title = element_text(hjust = 0, vjust = 1, family = "Times", size = 16), 
        axis.title.x = element_text(family = "Times", colour = "grey70", size = 10, hjust = 0))



```


#### The command below is helpful for debugging, please don't change it

```{r echo=FALSE}
sessionInfo()
```


